---
title: "UniFrac Calculations"
---

```{css, echo=FALSE}

.sourceCode           { display: none;   }
math[display="block"] { font-size: 22px; }

img {
  background-color: #fff;
  border: 2px gray solid;
}

.center {
  margin-left: auto;
  margin-right: auto;
}

.otu_matrix {
  width: 80%; 
  border-width: 0;
  border-collapse: separate;
  padding: 10px;
  font-size: 14pt;
  margin-left: auto;
  margin-right: auto;
}

.def_table {
  font-size: 12pt;
  margin-left: auto;
  margin-right: auto;
}
.def_table td {
  padding: 10px; 
}

.level2 {
  margin-top: 40px;
}

.level1 {
  margin-top: 50px;
}

```



```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  dev      = 'ragg_png' )
```


# Introduction

The different UniFrac algorithms are listed below, along with examples for calculating them.


<br>

# Input Data

```{r input_data}

mtx <- matrix(
  data     = c(0, 0, 9, 3, 3, 1, 4, 2, 8, 0), 
  nrow     = 5, 
  dimnames = list(paste0('OTU_', 1:5), paste0('Sample_', 1:2)) )

phy <- fastbiom::read_tree(
  newick      = "(((OTU_1:0.8,OTU_2:0.5):0.4,OTU_3:0.9):0.2,(OTU_4:0.7,OTU_5:0.3):0.6);",
  underscores = TRUE )

L <- phy$edge.length
A <- c(9,0,0,0,9,6,3,3)
B <- c(7,5,1,4,2,8,8,0)
```

* An OTU matrix with two samples and five OTUs.
* A phylogenetic tree for those five OTUs.

<div style='text-align:center'>

```{r input_data_mtx, results='asis'}
knitr::kable(t(mtx), format="html", table.attr='class="otu_matrix" cellspacing="0"', align='c')
```

<small><i>Shown transposed. Fastbiom expects samples in columns and OTUs in rows.</i></small>
</div>

<br>

```{r input_data_tree, fig.asp = 0.5, fig.align = 'center', out.width = "75%", fig.width = 6}
par(xpd = NA)
ape::plot.phylo(
  x          = phy, 
  direction  = 'downwards', 
  srt        = 90, 
  adj        = 0.5, 
  no.margin  = TRUE,
  underscore = TRUE,
  x.lim      = c(0.5, 5.5) )

ape::edgelabels(phy$edge.length, bg = 'white', frame = 'none', adj = -0.4)
```


<br>

# Definitions

The branch indices (green circles) are used for ordering the $L$, $A$, and $B$ arrays. Values for $L$ are drawn from the input phylogenetic tree. Values for $A$ and $B$ are the total number of OTU observations descending from that branch; $A$ for Sample_1, and $B$ for Sample_2.

```{r definitions, fig.asp = 0.5, fig.align = 'center', out.width = "75%", fig.width = 6}
local({
  
  phy$edge.length <- c(1, 1, 1, 1, 2, 1, 2, 2)
  
  par(xpd = NA)
  ape::plot.phylo(
    x               = phy, 
    direction       = 'downwards', 
    srt             = 90, 
    adj             = 0.5, 
    no.margin       = TRUE,
    underscore      = TRUE,
    x.lim           = c(.8, 6) )
  
  ape::edgelabels(1:8, frame = 'circle')
  
  ape::edgelabels(paste('A =', A), bg = 'white', frame = 'none', adj = c(-0.4, -1.2))
  ape::edgelabels(paste('B =', B), bg = 'white', frame = 'none', adj = c(-0.4,  0.0))
  ape::edgelabels(paste('L =', L), bg = 'white', frame = 'none', adj = c(-0.3,  1.2))
})
```

<br>

<div style='text-align:center'>
  <table class='def_table'>
  <tr><td> $n = 8$         </td><td> Number of branches               </td></tr>
  <tr><td> $A = \{`r A`\}$ </td><td> Branch **weights** for Sample_1. </td></tr>
  <tr><td> $B = \{`r B`\}$ </td><td> Branch **weights** for Sample_2. </td></tr>
  <tr><td> $A_T = 15$      </td><td> Total OTU counts for Sample_1.   </td></tr>
  <tr><td> $B_T = 15$      </td><td> Total OTU counts for Sample_2.   </td></tr>
  <tr><td> $L = \{`r L`\}$ </td><td> The branch **lengths**.          </td></tr>
  </table>
</div>

<br>


# Unweighted

* [Lozupone et al, 2005](https://doi.org/10.1128/AEM.71.12.8228-8235.2005): Unweighted UniFrac
* R Package [fastbiom](https://cran.r-project.org/package=fastbiom): `unweighted_unifrac()`
* R Package [phyloseq](https://doi.org/doi:10.18129/B9.bioc.phyloseq): `UniFrac(weighted=FALSE)`
* R Package [abdiv](https://doi.org/10.32614/CRAN.package.abdiv): `unweighted_unifrac()`
* [qiime2](https://qiime2.org/) `qiime diversity beta-phylogenetic --p-metric unweighted_unifrac`
* [mothur](https://mothur.org/): `unifrac.unweighted()`


First, transform A and B into presence (1) and absence (0) indicators.

<table class='center' style='width:75%'><tr><td style='border-width:0'>

\begin{align*}
A  &= \{`r A`\} \\
A' &= \{`r as.numeric(A > 0)`\}
\end{align*}

</td><td style='border-width:0'>

\begin{align*}
B  &= \{`r B`\} \\
B' &= \{`r as.numeric(B > 0)`\}
\end{align*}

</td></tr></table>


Then apply the formula:

\begin{align*}
U &= \displaystyle \frac{\sum_{i = 1}^{n} L_i(|A'_i - B'_i|)}{\sum_{i = 1}^{n} L_i(max(A'_i,B'_i))}  \\ \\
U &= \displaystyle \frac{L_1(|A'_1-B'_1|) + L_2(|A'_2-B'_2|) + \cdots + L_n(|A'_n-B'_n|)}{L_1(max(A'_1,B'_1)) + L_2(max(A'_2,B'_2)) + \cdots + L_n(max(A'_n,B'_n))}  \\ \\
U &= \displaystyle \frac{0.2(|1-1|) + 0.4(|0-1|) + \cdots + 0.3(|1-0|)}{0.2(max(1,1)) + 0.4(max(0,1)) + \cdots + 0.3(max(1,0))}  \\ \\
U &= \displaystyle \frac{0.2(0) + 0.4(1) + 0.8(1) + 0.5(1) + 0.9(0) + 0.6(0) + 0.7(0) + 0.3(1)}{0.2(1) + 0.4(1) + 0.8(1) + 0.5(1) + 0.9(1) + 0.6(1) + 0.7(1) + 0.3(1)}  \\ \\
U &= \displaystyle \frac{0.4 + 0.8 + 0.5 + 0.3}{0.2 + 0.4 + 0.8 + 0.5 + 0.9 + 0.6 + 0.7 + 0.3}  \\ \\
U &= \displaystyle \frac{2}{4.4}  \\ \\
U &= 0.4545455
\end{align*}


# Weighted

* [Lozupone et al, 2007](https://doi.org/10.1128/AEM.01996-06): Raw Weighted UniFrac
* R Package [fastbiom](https://cran.r-project.org/package=fastbiom): `weighted_unifrac()`
* R Package [phyloseq](https://doi.org/doi:10.18129/B9.bioc.phyloseq): `UniFrac(weighted=TRUE, normalized=FALSE)`
* R Package [abdiv](https://doi.org/10.32614/CRAN.package.abdiv): `weighted_unifrac()`
* [qiime2](https://qiime2.org/) `qiime diversity beta-phylogenetic --p-metric weighted_unifrac`

\begin{align*}
W &= \sum_{i = 1}^{n} L_i|\frac{A_i}{A_T} - \frac{B_i}{B_T}|  \\ \\
W &= L_1|\frac{A_1}{A_T} - \frac{B_1}{B_T}| + L_2|\frac{A_2}{A_T} - \frac{B_2}{B_T}| + \cdots + L_n|\frac{A_n}{A_T} - \frac{B_n}{B_T}|  \\ \\
W &= 0.2|\frac{9}{15} - \frac{7}{15}| + 0.4|\frac{0}{15} - \frac{5}{15}| + \cdots + 0.3|\frac{3}{15} - \frac{0}{15}|  \\ \\
W &= 0.02\overline{6} + 0.1\overline{3} + 0.05\overline{3} + 0.1\overline{3} + 0.42 + 0.08 + 0.2\overline{3} + 0.06  \\ \\
W &= 1.14
\end{align*}


# Normalized

* [Lozupone et al, 2007](https://doi.org/10.1128/AEM.01996-06): Normalized Weighted UniFrac
* R Package [fastbiom](https://cran.r-project.org/package=fastbiom): `normalized_unifrac()`
* R Package [phyloseq](https://doi.org/doi:10.18129/B9.bioc.phyloseq): `UniFrac(weighted=TRUE, normalized=TRUE)`
* R Package [abdiv](https://doi.org/10.32614/CRAN.package.abdiv): `weighted_normalized_unifrac()`
* [qiime2](https://qiime2.org/) `qiime diversity beta-phylogenetic --p-metric weighted_normalized_unifrac`
* [mothur](https://mothur.org/): `unifrac.weighted()`

\begin{align*}

N &= \displaystyle \frac
  {\sum_{i = 1}^{n} L_i|\frac{A_i}{A_T} - \frac{B_i}{B_T}|}
  {\sum_{i = 1}^{n} L_i(\frac{A_i}{A_T} + \frac{B_i}{B_T})}
  \\ \\
  
N &= \displaystyle \frac
  {L_1|\frac{A_1}{A_T} - \frac{B_1}{B_T}| + L_2|\frac{A_2}{A_T} - \frac{B_2}{B_T}| + \cdots + L_n|\frac{A_n}{A_T} - \frac{B_n}{B_T}|}
  {L_1(\frac{A_1}{A_T} + \frac{B_1}{B_T}) + L_2(\frac{A_2}{A_T} + \frac{B_2}{B_T}) + \cdots + L_n(\frac{A_n}{A_T} + \frac{B_n}{B_T})}
  \\ \\
  
N &= \displaystyle \frac
  {0.2|\frac{9}{15} - \frac{7}{15}| + 0.4|\frac{0}{15} - \frac{5}{15}| + \cdots + 0.3|\frac{3}{15} - \frac{0}{15}|}
  {0.2(\frac{9}{15} + \frac{7}{15}) + 0.4(\frac{0}{15} + \frac{5}{15}) + \cdots + 0.3(\frac{3}{15} + \frac{0}{15})}
  \\ \\
  
N &= \displaystyle \frac
  {0.02\overline{6} + 0.1\overline{3} + 0.05\overline{3} + 0.1\overline{3} + 0.42 + 0.08 + 0.2\overline{3} + 0.06}
  {0.21\overline{3} + 0.1\overline{3} + 0.05\overline{3} + 0.1\overline{3} + 0.66 + 0.56 + 0.51\overline{3} + 0.06}
  \\ \\
  
N &= \displaystyle \frac{1.14}{2.326667}  \\ \\
N &= 0.4899713
\end{align*}



# Generalized

* [Chen et al. 2012](https://doi.org/10.1093/bioinformatics/bts342): Generalized UniFrac
* R Package [fastbiom](https://cran.r-project.org/package=fastbiom): `generalized_unifrac(alpha = 0.5)`
* R Package [GUniFrac](https://cran.r-project.org/web/packages/GUniFrac/index.html): `GUniFrac(alpha = 0.5)`
* R Package [abdiv](https://doi.org/10.32614/CRAN.package.abdiv): `generalized_unifrac(alpha = 0.5)`
* [qiime2](https://qiime2.org/) `qiime diversity beta-phylogenetic --p-metric generalized_unifrac -a 0.5`

\begin{align*}

G &= \displaystyle \frac
  {\sum_{i = 1}^{n} L_i(\frac{A_i}{A_T} + \frac{B_i}{B_T})^{\alpha}
    |\displaystyle \frac
      {\frac{A_i}{A_T} - \frac{B_i}{B_T}}
      {\frac{A_i}{A_T} + \frac{B_i}{B_T}} |}
  {\sum_{i = 1}^{n} L_i(\frac{A_i}{A_T} + \frac{B_i}{B_T})^{\alpha}}  \\ \\

G &= \displaystyle \frac
  {
    L_1(\frac{A_1}{A_T} + \frac{B_1}{B_T})^{0.5}
      |\displaystyle \frac
        {\frac{A_1}{A_T} - \frac{B_1}{B_T}}
        {\frac{A_1}{A_T} + \frac{B_1}{B_T}}| + 
    \cdots + 
    L_n(\frac{A_n}{A_T} + \frac{B_n}{B_T})^{0.5}
      |\displaystyle \frac
        {\frac{A_n}{A_T} - \frac{B_n}{B_T}}
        {\frac{A_n}{A_T} + \frac{B_n}{B_T}}|
  }{
    L_1(\frac{A_1}{A_T} + \frac{B_1}{B_T})^{0.5} + 
    \cdots + 
    L_n(\frac{A_n}{A_T} + \frac{B_n}{B_T})^{0.5}
  }  
  \\ \\

G &= \displaystyle \frac
  {
    0.2(\frac{9}{15} + \frac{7}{15})^{0.5}
      |\displaystyle \frac
        {\frac{9}{15} - \frac{7}{15}}
        {\frac{9}{15} + \frac{7}{15}}| + 
    \cdots + 
    0.3(\frac{3}{15} + \frac{0}{15})^{0.5}
      |\displaystyle \frac
        {\frac{3}{15} - \frac{0}{15}}
        {\frac{3}{15} + \frac{0}{15}}|
  }{
    0.2(\frac{9}{15} + \frac{7}{15})^{0.5} + 
    \cdots + 
    0.3(\frac{3}{15} + \frac{0}{15})^{0.5}
  }  
  \\ \\
  
G &\approx \displaystyle \frac
  {0.03 + 0.23 + 0.21 + 0.26 + 0.49 + 0.08 + 0.27 + 0.13}
  {0.21 + 0.23 + 0.21 + 0.26 + 0.77 + 0.58 + 0.60 + 0.13}
  \\ \\
  
G &= \displaystyle \frac{1.701419}{2.986235}  \\ \\
G &= 0.569754

\end{align*}



# Variance Adjusted

* [Chang et al, 2011](https://doi.org/10.1186/1471-2105-12-118): Variance Adjusted Weighted (VAW) UniFrac
* R Package [fastbiom](https://cran.r-project.org/package=fastbiom): `variance_adjusted_unifrac()`
* R Package [abdiv](https://doi.org/10.32614/CRAN.package.abdiv): `variance_adjusted_unifrac()`
* [qiime2](https://qiime2.org/) `qiime diversity beta-phylogenetic --p-metric weighted_normalized_unifrac --p-variance-adjusted`

\begin{align*}

V &= \displaystyle \frac
  {\sum_{i = 1}^{n} L_i\displaystyle \frac
    {|\frac{A_i}{A_T} - \frac{B_i}{B_T}|}
    {\sqrt{(A_i + B_i)(A_T + B_T - A_i - B_i)}} }
  {\sum_{i = 1}^{n} L_i\displaystyle \frac
    {\frac{A_i}{A_T} + \frac{B_i}{B_T}}
    {\sqrt{(A_i + B_i)(A_T + B_T - A_i - B_i)}} }  \\ \\

V &= \displaystyle \frac
  {
    L_1\displaystyle \frac
    {|\frac{A_1}{A_T} - \frac{B_1}{B_T}|}
    {\sqrt{(A_1 + B_1)(A_T + B_T - A_1 - B_1)}} + 
    \cdots + 
    L_n\displaystyle \frac
    {|\frac{A_n}{A_T} - \frac{B_n}{B_T}|}
    {\sqrt{(A_n + B_n)(A_T + B_T - A_n - B_n)}}
  }{
    L_1\displaystyle \frac
    {\frac{A_1}{A_T} + \frac{B_1}{B_T}}
    {\sqrt{(A_1 + B_1)(A_T + B_T - A_1 - B_1)}} + 
    \cdots + 
    L_n\displaystyle \frac
    {\frac{A_n}{A_T} + \frac{B_n}{B_T}}
    {\sqrt{(A_n + B_n)(A_T + B_T - A_n - B_n)}}
  }  
  \\ \\

V &= \displaystyle \frac
  {
    0.2\displaystyle \frac
    {|\frac{9}{15} - \frac{7}{15}|}
    {\sqrt{(9 + 7)(15 + 15 - 9 - 7)}} + 
    \cdots + 
    0.3\displaystyle \frac
    {|\frac{3}{15} - \frac{0}{15}|}
    {\sqrt{(3 + 0)(15 + 15 - 3 - 0)}}
  }{
    0.2\displaystyle \frac
    {\frac{9}{15} + \frac{7}{15}}
    {\sqrt{(9 + 7)(15 + 15 - 9 - 7)}} + 
    \cdots + 
    0.3\displaystyle \frac
    {\frac{3}{15} + \frac{0}{15}}
    {\sqrt{(3 + 0)(15 + 15 - 3 - 0)}}
  }  
  \\ \\
  
V &\approx \displaystyle \frac
  {0.002 + 0.012 + 0.010 + 0.013 + 0.029 + 0.005 + 0.016 + 0.007}
  {0.014 + 0.012 + 0.010 + 0.013 + 0.046 + 0.037 + 0.036 + 0.007}
  \\ \\
  
V &= \displaystyle \frac{4.09389}{4.174402}  \\ \\
V &= 0.9807128

\end{align*}


